---
title: "Code along 01 - Himalayan Mountain Expeditions"
author: "Laurie Baker"
date: "09/09/2021"
output: html_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(skimr)
```

The data come from [TidyTuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/readme.md).
TidyTuesday is a weekly social data project for the R community.
Read more about TidyTuesday [here](https://github.com/rfordatascience/tidytuesday) and see people's contributions on Twitter under the [\#tidytuesday hashtag](https://twitter.com/search?q=tidytuesday&src=typed_query).

```{r load-data, message = FALSE}
members     <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/members.csv")
expeditions <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv")
peaks       <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/peaks.csv")
```

## Data overview

### Members

```{r}
skim(members)
```

### Peaks

```{r}
skim(peaks)
```

## Popular peaks

```{r}
members %>%
  count(peak_name) %>%
  arrange(desc(n)) %>%
  mutate(proportion = n / sum(n))
```

## Hired and died

```{r}
members %>%
  count(hired, died) %>%
  group_by(hired) %>%
  mutate(proportion = n / sum(n))
```

```{r}
ggplot(members, aes(x = hired, fill = died)) +
  geom_bar(position = "fill")
```

Go to RStudio Cloud and try it yourself in the project titled `Code Along 01 - Himalayan Climbing Expeditions`.
