---
title: "Taking your plot to the next level - themes, axes, and annotation"
subtitle: "<br><br> College of the Atlantic"
author: "Laurie Baker"
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggthemes)
library(gapminder)
```

---
# Recap: How do we express visuals in words?


- **Data** to be visualized


--

- **`Aes`thetic mappings** from data to visual component


--

- **`Geom`etric objects** that appear on the plot


--

- **`Stat`istics** transform data on the way to visualization


--

- **`Coord`inates** organize location of geometric objects


--

- **`Scale`s** define the range of values for aesthetics


--

- **`Facet`s** group into subplots

--
- **`Theme`s** the visual elements of the plot not linked to the data

---
# Today's Focus

- **Theme**s: the visual elements of the plot not linked to the data
- **Geom**etric objects: `geom_vline` and `geom_hline`
- **Scale**s: define the range of values for aesthetics
- **Annotations:** text, highlight, etc. 
---
## Packages

```{r, eval=FALSE}
library(tidyverse)
library(ggthemes)
library(gapminder)
```

---
# head(gapminder)

```{r head-gapminder, echo=FALSE}
head(gapminder)
```

---
# glimpse(gapminder)

```{r summary-gapminder, echo=FALSE, comment=""}
glimpse(gapminder)
```

---
# Life Expectancy

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}
(p1 <- gapminder |>
  filter(continent == "Asia") |>
ggplot(aes(x = year,
      y = lifeExp, 
      group = country)) +
  geom_line() +
  labs(x = "Year",
       y = "Life Expectancy (years)",
       title = "Life Expectancy in Asia"))
```
]
---
# themes - black and white

```{r}
p1 + theme_bw()
```

---
# themes - dark

```{r}
p1 + theme_dark()
```

---
# themes - minimal

```{r}
p1 + theme_minimal()
```

---
# themes - build your own

.right-column[
Huge number of parameters, grouped by plot area:

- Global options: `line`, `rect`, `text`, `title`
- `axis`: x-, y- or other axis title, ticks, lines
- `legend`: Plot legends
- `panel`: Actual plot area
- `plot`: Whole image
- `strip`: Facet labels
]
---
# themes - build your own

.right-column[
Theme options are supported by helper functions:

- `element_blank()` removes the element
- `element_line()`
- `element_rect()`
- `element_text()`

```{r}
my_theme <- theme_bw() + theme(text = element_text(color = "hotpink", size = 20), axis.ticks.x = element_blank())
```
```

]

---

.right-column[

```{r, eval = TRUE, out.width="90%", fig.width=6}
p1 + my_theme
```
]

---
# ggthemes - Economist

```{r}
p1 + theme_economist()
```

---
# Life Expectancy - highlight
.panelset[
```{r panelset = c(output = "Plot", source = "Code")}
(p1 <- p1 +
  geom_line(data = gapminder |> filter(country == "Cambodia"),
  color = "orange", 
  lwd = 2, 
  aes(x = year, 
  y = lifeExp, 
  group = country)) +
   theme_bw())
```
]
- [Cambodian Genocide](https://sfi.usc.edu/collections/cambodian-genocide)

---
# Add annotations with `geom_vline`

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}
(p1 <- p1 +
  geom_vline(xintercept = 1975, linetype = "dotted", color = "red") +
  geom_vline(xintercept = 1979, linetype = "dotted", color = "red"))
```
]

---
# Add annotations with `annotate`

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}
p1 +
  annotate("rect", 
  xmin = 1975, 
  xmax = 1979, 
  ymin = 0, 
  ymax = 100, 
  alpha = 0.2, 
  fill = "red")
```
]
---
# Scales

.left-column[

```r
+ scale_*_*()
```
]

.right-column[ 
`scale` + `_` + `<aes>` + `_` + `<type>` + `()`

`<aes>` = parameter to adjust; `<type>` = Parameter Type
]

---
# Scales

.left-column[
### Scales

```r
+ scale_*_*()
```
]

.right-column[
- I want to use a different color palette<br>`scale_fill_discrete()`<br>`scale_color_continuous()`
]

---
# Scales

- I want to rescale y-axis as log<br>`scale_y_log10()`

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}
p1 +
  annotate("rect", xmin = 1975, xmax = 1979, ymin = 0, ymax = 100, alpha = 0.2, fill = "red") +
  scale_y_log10()
```

]


---
# Scales

- I want to change my continuous x-axis<br>`scale_x_continuous()`

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}
p1 +
  annotate("rect", xmin = 1975, xmax = 1979, ymin = 0, ymax = 100, alpha = 0.2, fill = "red") +
  scale_x_continuous(breaks = seq(from = 1952, to = 2007, by = 5))
```

]


---
# Annotation - geom_text

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}
p1 +
  annotate("rect", 
  xmin = 1975, 
  xmax = 1979, 
  ymin = 0, 
  ymax = 100, 
  alpha = 0.2, 
  fill = "red") +
  geom_text(aes(x = 1985, y = 25), 
  label = "Khmer Rouge")
  
```
]
---
# Annotation - geom_text

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}

(p2 <- gapminder |>
  filter(country == "Portugal") |>
ggplot(aes(x = year,
      y = lifeExp)) +
  geom_point() +
  labs(x = "Year",
       y = "Life Expectancy (years)")) +
  geom_text(aes(label = year))
  
```

]


---
# Annotation - geom_text

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}

(p2 <- gapminder |>
  filter(country == "Portugal") |>
ggplot(aes(x = year,
      y = lifeExp)) +
  geom_point() +
  labs(x = "Year",
       y = "Life Expectancy (years)",
       title = "Life Expectancy in Portugal over time") +
  geom_text(aes(label = year), nudge_x = 3.75, nudge_y = -0.3))
  
```

]


---
# Annotation - gghighlight

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}

gapminder |>
ggplot(aes(x = year,
      y = lifeExp, 
      group = country, 
      color = continent)) +
  geom_line() +
  labs(x = "Year",
       y = "Life Expectancy (years)") +
  gghighlight(mean(lifeExp) > 75) 
```
]
---
# Annotation - gghighlight

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}

gapminder |>
ggplot(aes(x = year,
      y = lifeExp, 
      group = country, 
      color = continent)) +
  geom_line() +
  labs(x = "Year",
       y = "Life Expectancy (years)") +
  facet_wrap(~continent) +
  gghighlight(mean(lifeExp) > 74.5) 
```
]

---
# Annotation - gghighlight

.panelset[
```{r panelset = c(output = "Plot", source = "Code")}

gapminder |>
ggplot(aes(x = year,
      y = lifeExp, 
      group = country, 
      color = continent)) +
  geom_line() +
  labs(x = "Year",
       y = "Life Expectancy (years)") +
  gghighlight(min(lifeExp) > 55 & max(lifeExp) < 70)
```
]

---
# Your Turn

- Create your own plot highlighting a county of interest and a labelled historical event. 
