---
title: "Looking for Bigfoot"
author: "Your Name Here"
output: github_document
---

```{r load-packages, message = FALSE}
library(visdat)
library(naniar)
library(UpSetR)
library(leaflet)
library(lubridate)
```

We will use the data on bigfoot sitings from tidytuesday

```{r}
bigfoot <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-09-13/bigfoot.csv')

```

```{r}
glimpse(bigfoot)
```

Let's focus on a few variables for now

```{r}
bigfoot <- bigfoot %>%
  select(county, state, observed, season, latitude, longitude, date, moon_phase, precip_intensity, visibility) %>%
  mutate(year = year(date))

```

## Create a plot to visualize the data types we have

```{r data-type viz}

```

## Create a plot that explores the amount of data that is missing

```{r percent missing}

```

Which variable is missing the highest percentage of data?

## Create a plot that explores the relationship in missing data between moon_phase and precipitation intensity

```{r}

```

## Create a plot that explores the relationship in missing data between moon_phase and precipitation intensity faceted by season

```{r}

```

## Which variables tend to be missing together? Check ?

```{r missing variables}

```

## Bonus: Create a leaflet map of Bigfoot sitings between 1940-1990

```{r leaflet, eval = FALSE}
bigfoot %>%
  filter(________) %>%
leaflet() %>%
  addProviderTiles(providers$OpenStreetMap) %>%
  addCircleMarkers(lng = _____, 
                   lat = _____, 
                   label = ~observed)

```
