---
title: "Visualizing Missing Data"
subtitle: "<br><br> College of the Atlantic"
author: "Laurie Baker"
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(visdat)
library(naniar)
library(UpSetR)
```

class: middle

# Visualizing Missing Data

---
## Origins of Missing Data

- What are the origins of missing data?

---
## Packages for visualizing missing data


```{r load-packages}
library(visdat)
library(naniar)
library(UpSetR)
```

---
## Inspecting the data types

Let's explore a data set of imaginary people who have been randomly generated.
```{r glimpse data}
glimpse(typical_data)
```
* Which of the variable's data types have likely been incorrectly identified?

.footnote[For more info about the randomly generated data run `?typical_data`]

---
## Summarizing Missing data

```{r summary data}
summary(typical_data)
```
* `summary` can give us some ideas of missing data

---
## Visualizing Data Types 

```{r vis data types}
vis_dat(typical_data)
```

---
## Visualizing Missing Data

```{r vis missing data}
vis_miss(typical_data)
```

---
## Calculating how much is missing

```{r missing calculations}
# From nanianr package:

n_miss(typical_data) # number missing

n_complete(typical_data) # number complete

pct_miss(typical_data) # percent missing
```

---
# Calculating how much is missing by column

```{r missing variable summary}
miss_var_summary(typical_data)
```

---
# Visualizing missing data by variable

```{r plot missing data by variable}
gg_miss_var(typical_data)
```

---
# Visualizing percent missing data by variable

```{r plot percent missing}
gg_miss_var(typical_data, show_pct = TRUE) +
  ylim(0, 100)
```

---
# Visualizing missing data by another variable

```{r plot missing by facet}
gg_miss_var(typical_data,
            facet = Race)
```

---
# Missingness across factors

```{r plot missing percent heatmap}
gg_miss_fct(x = typical_data, fct = Race) +
  labs(y = "Variable")
```


---
# Missingness across cases

```{r plot missing cases}
gg_miss_case(typical_data)
```


---
# Missingness across cases

```{r understanding missing case frequency}

typical_data %>%
  as_shadow_upset() %>%
  upset()
```

---
# Daily Air Quality Data in NYC, May-Sept 1973

```{r visualizing data types of airquality}
vis_dat(airquality)
```

---
# Relationships in missing values

Returning to the `ggplot` warnings

```{r scatterplot of Solar radiation and Ozone}
ggplot(airquality, 
       aes(x = Solar.R, y = Ozone)) + 
  geom_point()
```


---
# Relationships in missing values

```{r plot of missing points}
airquality %>%
  ggplot(aes(x = Ozone, y = Solar.R)) +
  geom_miss_point()

```

---
# Relationships in missing values

```{r plot missing relationships with facets}
airquality %>%
  ggplot(aes(x = Ozone, y = Solar.R)) +
  geom_miss_point() +
  facet_wrap(~Month)

```

---
# Visualizing relationships in missing variables

```{r visualizing relationships in missing variables}
airquality %>%
    bind_shadow() %>%
    ggplot(aes(x = Solar.R, fill = Ozone_NA)) +
  geom_histogram(binwidth = 20)
```


---
# Consequences of Missing Data

- So you have data that is missing? What now?

---
# Your Turn


```{r, fig.alt = "Outline of Sasquatch", fig.height=4, echo=FALSE}
knitr::include_graphics("img/Sasquatch.svg")
```

Your Turn:

<a href="https://commons.wikimedia.org/wiki/File:Sasquatch.svg">Happybluemo</a>, <a href="https://creativecommons.org/licenses/by-sa/4.0">CC BY-SA 4.0</a>, via Wikimedia Commons

---
# Thank you