---
title: "Working with regular expressions"
subtitle: "<br><br> College of the Atlantic"
author: "Laurie Baker"
output:
  xaringan::moon_reader:
    css: ["../xaringan-themer.css", "../slides.css"]
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      ratio: "16:9"
      highlightLines: true
      highlightStyle: solarized-light
      countIncrementalSlides: false
---

```{r child = "../setup.Rmd"}
```

```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext)
```

class: middle

# Regular Expressions

---
## Regular Expressions

Pattern matching functions use patterns, otherwise known as "regular expressions" or "regex", to identify specific characteristics in strings.

- Reference material: [R4DS Regular Expressions](https://r4ds.hadley.nz/regexps)

---
## Pattern basics

- We can use `str_view()` to show the elements of the string vector that match the regular expression `<>`.

```{r}
library(stringr)
library(babynames)
```

The simplest patterns consist of letters and numbers which match those characters:

```{r}
str_view(string = fruit, pattern = "berry")
```

*Note that the first argument in `str_view` is for the `string` and the second argument is the `pattern`. 

---
## Literal characters and meta characters

- **Literal characters** Letters and numbers
- **Meta characters** Includes most punctuation characters like `.`, `+`, `*`, `[`, `]`, and `?` which have special meanings. 
    - For example, `.` will match any character. So `"a." will match any string that contains an `"a"` followed by another character:
    
```{r}
str_view(c("apple", "orange", "pomegranate"), pattern = "a.")
```

We could find names that include an `"a"` followed by three letters, followed by an `"e"`. 

```{r}
baby1880 <- babynames %>%
  filter(year == 1880)

str_view(baby1880$name, pattern = "a...e")
```

---
## Quantifiers

**Quantifiers** control how many times a pattern can match.

- `?` makes a pattern optional (i.e. it matches 0 or 1 times)
- `+` lets a pattern repeat (i.e. matches at least once)
- `*` lets a pattern be optional or repeat (i.e. it matches any number of times, including 0)

```{r}
str_view(c("a", "ab", "abb"), pattern = "ab?")
```

```{r}
str_view(c("a", "ab", "abb"), pattern = "ab+")
```

```{r}
str_view(c("a", "ab", "abb"), pattern = "ab*")
```

- **Your Turn**: With your partner write a short description of what the pattern matches. 

---
## Character Classes

**Character classes** are defined by `[]` and let you match a set of characters

- e.g. `[lmno]` matches "l", "m", "n", or "o".
- starting with `^` will invert the match so that you can match anything **except** "l", "m", "n", or "o". 

```{r}
str_view(words, "[aeiou]x[aeiou]")

str_view(words, "[^aeiou]y[^aeiou]")
```

---
## Your Turn

Work on Exercises 1 and 2.


---
## Alternation

**alternation**, `|` can be used to pick between one or more alternative patterns.

- Fruits with "nut", "berry", or "apple".

```{r}
str_view(fruit, "apple|nut|berry")
```


- Names with a repeating vowel

```{r}
str_view(baby1880$name, pattern = "aa|ee|ii|oo|uu")
```

---
## Your Turn

Work on Exercises 3 and 4.


---
## Key stringr functions

- `str_detect` detect the presence or absence of a match
- `str_count` count how many matches are in each string
- `str_extract` extract the matches

---
## Detect matches

`str_detect()` returns a logical vector that is `TRUE` if the pattern matches an element of the character vector and `FALSE` otherwise.

```{r}
str_detect(c("a", "b", "c"), "[aeiou]")
```

`str_detect()` works well with `filter`

```{r}
baby1880 |>
  filter(str_detect(name, pattern = "x")) |>
  count(name, wt = n, sort = TRUE)
```

---
## Count matches

`str_count()` can be used to count the number of times a pattern occurs in a string.

```{r}
str_count("ababbaba", pattern = "ab")
str_view("ababbaba", pattern = "ab")
```

`str_count()` can be easily paired with `mutate()`

```{r}
# number of vowels and consonants for babynames with an x

babynames |>
  filter(str_detect(name, pattern = "x")) |>
  count(name) |>
  mutate(vowels = str_count(name, "[aeiou]"),
         consonants = str_count(name, "[^aeiou]")
         ) %>%
  arrange(vowels, consonants, n)
```

There's something off with our calculations. What is it?

---
## Regular expressions are case_sensitive

Options: 
- Add uppercase vowels: `str_count(name, pattern = "[aeiouAEIOU]")`
- Tell the regular expression to ignore case:
`str_count(name, regex("[aeiou]", ignore_case = TRUE))`
- Use str_to_lower() to convert the names to lower case:
`str_count(str_to_lower(name), "[aeiou]")`

---
## Replace values

- We can also modify matched with `str_replace()` and `str_replace_all()`. `str_replace` replaces the first match and `str_replace_all()` replaces all matches.

```{r}
babynames |>
  count(name) |>
  filter(str_detect(name, pattern = "[cC]")) |>
  mutate(new_name = str_replace_all(name, "[cC]", "cup"))
```

---
## Replace values

```{r small = T}
str_replace(fruit, pattern = "a", replacement = "e") # only the first instance
str_replace_all(fruit, pattern = "a", replacement = "e") # every instance
```

---
## Your Turn

Work on exercises 5 and 6
    
---
## Extracting Patterns

- You can extract pattern matches using `str_extract()`

```{r}
lab_fees <- c("100 dollars", "10$", "1500 USD")

str_extract(lab_fees, pattern = "[0-9]*")

```

---
## Extract phone numbers case study

#### **Maine phone numbers:**
    - Maine phone numbers begin with 207 and are followed by 7 more digits.
    - Regex expression: "^207[0-9]{3}[0-9]{4}"

```{r}
phone_numbers <- c("207-846-3630", "207-865-3823", "603-712-5043", "0783-792-9863", "+4478438971066", "207-865-7916")

str_extract(phone_numbers, pattern = "^207-[0-9]{3}-[0-9]{4}")
```

*N.B. `{3}` will repeat your regular expression pattern a specified number of times.

---
## Pattern details

- **Escaping** how to match metacharacters. 
- **Anchors** match the start or end of a string.
- More on **character classes** and their shortcuts.
- More on **quantifiers** which control how many times a pattern can match.

---
## Escaping

- Regular expressions use the backslash for escaping metacharacters.
- To match a ., we need the regexp `\.`. 
- Because we use strings to represent the regular expression, and `\` is also used as an escape symbol in strings, to create the regular expression `\.` we need the string `"\\."`

```{r Escaping}
str_view(c("abc", "a.c", "bef"), "a\\.c")
```


---
## Anchors

By default regular expressions will match any part of a string. To match the start or end you need to **anchor** the regular expression:
- `^` matches the start
- `$` to match the end.

```{r}
str_view(baby1880$name, "^G")
str_view(baby1880$name, "g$")
```

- `\b` can match the boundary between words

```{r}
str_view(fruit, "\\bapple\\b")
str_view(fruit, "apple")
```

---
## Your Turn

Try Exercise 7

---
## Character classes

You can construct your own **character class** or **set** with `[]`.
- `[abc]` matches "a", "b", or "c" and `[^abc]` matches any character except "a", "b", or "c". 
- `-` defines a range, e.g., `[a-z]` matches any lower case letter and `[0-9]` matches any number.
- `\` escapes special characters, so `[\^\-\.]` matches `^`, `-`, or `.`

```{r}
x <- "abcd ABCD 12345 -!@#%."

str_view(x, "[abc]+")

str_view(x, "[a-z0-9]+")

str_view(x, "[^a-z0-9]+")

```

---
## Character shortcuts

`.` matches any character apart from a newline. There are three other useful pairs:

- `\d` matches any digit; `\D` matches anything that **isn't** a digit
- `\s` matches any whitespace (e.g. space, tab, newline)' `\S` matches anything that isn't whitespace.
- `\w` matches any "word" character, i.e. letters and numbers; `\W` matches any "non-word" character.

```{r}
x <- "abcd ABCD 12345 -!@#%."

str_view(x, "\\d+")
str_view(x, "\\D+")
str_view(x, "\\s+")
str_view(x, "\\S+")
str_view(x, "\\w+")
str_view(x, "\\W+")
```

---
## Quantifiers

- **Quantifiers** control how many times a pattern matches. 
    - `?` (0 or 1 matches)
    - `+` (1 or more matches)
    - `*` (0 or more matches)

```{r}
spelling_bee <- c("color", "colour", "coloor")

str_view(spelling_bee, pattern = "colou?r")

lab_fees <- c("100 dollars", "10$", "1500 USD")
# \\d+ for one or more digits
str_extract(lab_fees, pattern = "\\d+")
```

---
## Number of matches

You can specify the number of matches precisely with `{}`

- `{n}` matches exactly n times
- `{n,}` matches **at least** n times
- `{n,m}` matches between n and m times

---
## Operator precedence and parentheses

- Does `ab+` match "a" followed by one or more "b"s, or does it match "ab" repeated any number of times?

--

- What does `^a|b$` match? Does it match the complete *string a* **OR** the complete *string b* or does it match a string starting with "a" or a string ending with "b"?

---
## Operator precedence and parentheses

- It's similar to PEMDAS or BEMDAS.
- Quantifiers have high precedence and alternation has low precedence which means `ab+` is equivalent to `a(b+)`, and `^a|b$` is equivalent to `(^a)|(b$)`.
- You can use parentheses to override the usual order or to match anything in the middle try `.+`.

```{r}

word_ex <- c("abba", "ababab", "arbitrary", "about", "aplomb")

str_view(word_ex, pattern = "ab+")
str_view(word_ex, pattern = "(ab)+")

str_view(word_ex, pattern = "^a|b$")
str_view(word_ex, pattern = "^a.+b$")


```


---
## Recap of Common expressions:
    - "a"  = is the letter "a"
    - "^a" = starts with the letter "a"
    - "a$" = ends with the letter "a"
    - "[ ]" = contains any letter (or number) within the brackets
    - "[ - ]" = contains any letter (or number) within this range
    - "[^ae]" = everything except these letters (or numbers)
    - "{3}" = repeats a regular expression.

For more expressions or examples, refer to <http://www.regular-expressions.info/refquick.html>


---
# Thank you